GEE models

### 2016-05-02 doc/manuscript.Rmd
Added Tony's edits to the manuscript

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
output:
    word_document:
        reference_docx: diabetes-care-template.docx
        fig_height: 6 
        fig_width: 5
        fig_caption: true
        keep_md: true
---

@@

Title: Total non-esterified fatty acids, but not composition, predict
progression in beta-cell dysfunction: The Prospective Metabolism and Islet Cell
Evaluation (PROMISE) cohort

@@

<!-- First, middle initial, last, highest academic degree -->
Author: Luke W. Johnston, Stewart Harris, Ravi Retnakaran, Bernard Zinman,
Zhen Liu, Richard P. Bazinet, Anthony J. Hanley

@@

@@

- Name: Anthony J. Hanley
- Current address:
- Phone number:
- Fax number:
- Email: anthony.hanley@utoronto.ca

Word count: 3527 / 4000

@@

<!-- 250 words -->
# Abstract

Objective:

Research Design and Methods:

Results:

Conclusions:

@@

TODO:
- Bootstrap for the longitudinal table data... like instead of an ANOVA/GEE for
the p-value...?
- What about adiponectin?
- Precise followup time for each participant?
- [X] Interaction by sex or ethnicity
- [X] Interaction by time
- [X] Sensitivity analysis with TG
- [X] PLS-DA with dysglycemia (for anyone who got it over the years).

@@

sfig_corr <- sfig('corr', 'Pearson correlation heatmap of non-esterified fatty acids (nmol/mL) and basic PROMISE participant characteristics for the baseline visit (2004-2006). Darkness of the colour indicates the size of the correlation, with blue indicating positive and orange indicating negative correlations.')
sfig_gee <- sfig('gee', 'Unadjusted generalized estimating equation modeling of the longitudinal association of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort. Only adjusted for time. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, presented as the dot size.')

@@

stab_gee_unadj <- stab('gee_unadj', 'Unadjusted generalized estimating equations models of the longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.  Adjusted for time only. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values were adjusted for the BH false discovery rate, indicated by asterisk.')
sfig_lcmm <- sfig('lcmm', 'Latent class mixed model analysis to identify individual classes of trajectories for log(ISSI-2) over the 6 years in the PROMISE cohort.')
sfig_plsda_groups <- sfig('plsda_groups', 'Clustering of extracted components from the partial least squares discriminant analysis on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis. {{colour}}, {{colour}}, {{colour}} lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis.')

@@

sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', 'Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis. {{colour}} lines indicate dysglycemia conversion or maintanence and {{colour}} lines indicate no dysglycemia status.')

@@

Total non-esterified fatty acids (NEFA) have been well documented to influence 
the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that 
exposure to high concentrations of NEFA can induce insulin resistance in
insulin-sensitive tissue such as the muscle or liver and impair beta-cell
production of insulin [@Rachek2014a;@Cnop2008a]. Epidemiological and clinical
studies have observed similar findings, showing that elevated total plasma NEFA
associates with risk for diabetes [@Daniele2014a;@Salgin2012a].

Much of the previous experimental work frequently use fatty acids such as
palmitic acid (16:0) or oleic acid (18:1n-9), or use a standard oil such as soybean
oil as the exposure and generalize as total NEFA. However, fatty acids comprise
multiple classes of molecules with diverse physiological functions, with few
studies comparing different fatty acids. For instance, one study suggests that
eicosapentaenoic acid (20:5n-3) can protect against the lipotoxic effect of
palmitic acid in the beta-cells [@Xiao2006].

Despite a sizable literature studying the role of NEFA in diabetes, there are 
important gaps in the research. In particular, there have been few long term 
studies in humans that have examined a range of NEFA species. Most studies have 
been on animals and cells, have been short term human infusion trials, or have 
been epidemiological studies that only looked at total NEFA. To date, there have
been no longitudinal studies examining the role of the composition of the NEFA 
blood lipid fraction on the pathogenesis of diabetes, which is key to 
understanding causal associations given the chronic nature of diabetes 
pathophysiology. Therefore, our research objective was to examine the role of
the serum NEFA composition on insulin sensitivity and beta-cell function in a
longitudinal cohort. Some specific hypotheses included that higher palmitic acid
while lower polyunsaturated fatty acids such as eicosapentaenoic acid would 
associate with lower insulin sensitivity and beta-cell function.

@@

Participants from London and Toronto, Canada, were recruited into the 
Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
for recruitment into PROMISE included having one or more risk factors for type 2
diabetes mellitus, such as obesity, hypertension, family history of diabetes, 
and/or a history of gestational diabetes or birth of a macrosomic infant. 
Participants aged 30 years and older (n=736) attended the baseline visit
between 2004-2006. Follow-up examinations occur every three years, with three
examination visits completed to date (2004-2006, 2007-2009, and 2010-2013).  The
current study used data on participants who did not have diabetes at baseline,
who returned for one or more follow-up examinations, and who had samples
available for fatty acid measurements (n=477). A diagram of the sample size at
each visit is shown in `r cite_sf('consort')`. At each examination,
participants undergo metabolic characterization, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics. Research ethics approval
was obtained from Mount Sinai Hospital and the University of Western Ontario. 
Data collection methods were standardized across the 2 centres and research
nurses were centrally trained.

@@

At each examination, an 8-12 hour fasting blood sample was draw from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and triglycerides (TG) were measured using Roche
Modular's enzymatic colometric tests (Mississauga, ON).  Both specific insulin
and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
Basel, Switzerland) immunoassay analyzer and electrochemiluminescence
immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
and the Des 31,32 circulating split form (Linco Res. Inc), and has a CV of 9.3%.
Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
Diagnostics) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
performed at the Banting and Best Diabetes Centre Core Lab at the University of
Toronto. Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and
diabetes were categorized using the 2006 WHO criteria [@WHO2006].

NEFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of heptadecanoic acid (17:0) {{confirm}} was added to the serum as an internal 
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction (NEFA, cholesteryl ester, phospholipid,
and triacylglycerol) was isolated using thin layer chromatography, which were
visualized under UV light after lightly spraying with 8-anilino-1-naphthalene
sulfonic acid (0.1% wt/vol) and then converted to fatty acid methyl esters with
14% boron trifluoride in methanol at 100°C for 1 h. Fatty acid methyl esters
were separated and quantified using a Varian-430 gas chromatograph (Varian, Lake
Forest, CA, USA) equipped with a Varian Factor Four capillary column and a flame
ionization detector, which were injected in splitless mode. Fatty acid
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a;
@Chen2011a].  There were 22 fatty acids measured in the NEFA fraction. Given the
diverse biology, as well as the complexity of the analyses, the other lipid
fractions were split into separate reports (see {{pl and ce paper}} for the
analysis of the phospholipid and cholesteryl ester fractions).

@@

Insulin sensitivity and beta-cell function indices were computed using the OGTT 
glucose and insulin. Insulin sensitivity was assessed using 1/HOMA-IR (1
divided by HOMA-IR) [@Matthews1985] and the Insulin Sensitivity Index (ISI)
[@Matsuda1999]. HOMA-IR largely reflects hepatic insulin resistance, while ISI
reflects whole-body insulin sensitivity [@AbdulGhani2007].  Beta-cell function
was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR)
and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
is a measure of the first phase insulin secretion while ISSI-2 is analogous to
the disposition index. Each index has been validated against gold standard
measures [@Matthews1985; @Matsuda1999; @Retnakaran2009].

@@

Height, weight, and waist circumference (WC) were measured at all clinic 
examinations using standard procedures.  WC was measured at the natural waist, 
defined as the narrowest part of the torso between the umbilicus and the xiphoid
process. BMI was calculated by dividing weight (kg) by height (m) squared.
Sociodemographic information, such as age, sex, and ethnicity, were determined
using questionnaires administered at each examination. In the lifestyle
questionnaire, physical activity was determined using a version of the
Modifiable Activity Questionnaire (MAQ) [@Kriska1990].  The MAQ collects 
information on leisure and occupational activity, including intensity, 
frequency, and duration, over the past year.  Each reported activity from the 
MAQ was weighted by its metabolic intensity allowing for the estimation of 
MET-hours per week.

@@

The primary outcome variables for the PROMISE cohort and this analyses were 
1/HOMA-IR, ISI, IGI/IR, and ISSI-2; outcome variables were log-transformed for
the statistical modeling. The primary predictor variables for this study were
the 22 NEFA as a mole percent (mol%) of the total fraction and as a 
concentration (nmol/mL). Pearson correlation coefficients were computed for the
individual NEFA and with other continuous variables.

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design 
an auto-regressive (AR1) correlation matrix was chosen for the GEE models, 
though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
is well suited to cohort studies given it's capacity to handle missed visits.
The predictor variables and continuous covariates were scaled (mean centered and
standardized). The NEFA variables were classified as *time-independent* (held
constant) as they were measured only at the baseline visit, while the outcome
variables and covariates were set as *time-dependent*. No imputation was
conducted on missing values.

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
(QIC). `r cite_st('qic')` shows the covariates compared using QIC and 
`r cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model 
we selected differed between insulin sensitivity and beta-cell function models. 
For insulin sensitivity, we adjusted for time, sex, ethnicity, baseline age, WC,
ALT, MET, alcohol intake, and family history of diabetes (M6). For beta-cell
function, the last few models were comparable in performance, so we chose the
simplest model (M8), adjusting for time, sex, ethnicity, baseline age, WC, ALT,
and family history of diabetes. After scaling, log-transforming, and
exponentiating the GEE estimates, the GEE results are interpreted as an expected
percent difference in the outcome variable for every SD increase in the
predictor variable given the covariates are held constant. While models with a
time interaction did not perform better than others for the QIC (M6.1 and M8.1)
in `r cite_st('qic')`, we ran models with a time by predictor term for each of
the NEFA to test if there was an interaction by time. Given that TG is a risk factor
for diabetes and that NEFA contribute to TG production, TG may act as a mediator
between NEFA and the outcomes. However, only a subset of the participants at the
6-yr visit had clinically measured TG. To determine the role of TG in the
association between NEFA and the outcomes, TG was included in the GEE model as a
sensitivity analysis. Lastly for the GEE models, we tested for an interaction
with sex or ethnicity by the predictor term for the outcome variables.

While GEE controls for the longitudinal design of the data, a limitation is that
it can not analyze the inherent multivariate nature of the composition of the 
NEFA fraction. As such, after running the GEE models for the longitudinal 
analysis, partial least squares discriminant analysis (PLS-DA) was used to
identify the patterns of NEFA composition against beta-cell function (using ISSI-2,
which has most of the associations in the GEE models) and for those who converted or
maintained dysglycemia status (either IFG, IGT, or DM) over the 6 years. Since
PLS-DA can't handle longitudinal data, the longitudinal data was included by
extracting latent trajectories of beta-cell function (ISSI-2) over the 6 years using
latent class mixed models (LCMM). A limitation of these methods, excluding LCMM,
is that they can not be adjusted for covariates.

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a], caret
`r packageVersion('caret')` for PLS-DA, and lcmm `r packageVersion('lcmm')` for
LCMM. The R code for this manuscript is available at {{link}}. Results were
statistically significance at p<0.05, after adjusting for multiple testing using
the Benjamini-Hochberg False Discovery Rate [@Benjamini1995a].

@@

gee_results_unadj <- bind_rows(
    analyze_gee(gee_df, outcomes, ne_pct, 'VN', 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE)
    ) %>% 
    tidy_gee_results()
```

```{r lcmmPls, cache=TRUE}
# include also the 3-yr and 6-yr data for training/testing.
lcmm_results <- analyze_lcmm(ds)
lcmm_data <- combine_lcmm_data(ds, lcmm_results)
plsda_results <- analyze_plsda(lcmm_data)
plsda_dysgly <- dysgly_data %>% 
    mutate(ConvertDysgly = factor(ConvertDysgly, levels = c(0, 1), labels = c('No', 'Yes'))) %>% 
    analyze_plsda('ConvertDysgly')
```

@@

Within this subset of the PROMISE cohort at the baseline visit, 
`r sex.npct$npct` were females and `r ethn.npct$npct` had European-ancestry.
The average age of the participants was `r aide::average(ds0yr$Age)` years 
(`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
subset of the cohort, `r fhd.npct$npct`, had a family history of diabetes.
The primary outcome variables had a significant median decline between 
`r percent.change.outcomes$chg` (p`r percent.change.outcomes$p`) over the 6
years in this subset of the PROMISE cohort.
Consistent with this decline, there were 42 (9%) <!--`r dm.convert`-->
participants who developed diabetes while `r predm.convert` participants either
converted to or maintained IFG or IGT status over the 6 years.

>>>>

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
output:
    word_document:
        reference_docx: diabetes-care-template.docx
        fig_height: 6 
        fig_width: 7
        fig_caption: true
        keep_md: true
---

@@

Title: Total non-esterified fatty acids, but not individual species, predict
progression in beta-cell dysfunction: The Prospective Metabolism and Islet Cell
Evaluation (PROMISE) cohort

@@

<!-- First, middle initial, last, highest academic degree -->
Author: Luke W. Johnston (1), MSc; Stewart B. Harris (2), MD; Ravi Retnakaran
(3,4), MD; Bernard Zinman (3,4), MD; Adria Giacca (5), PhD; Zhen Liu (1), PhD;
Richard P. Bazinet (1), PhD; and Anthony J. Hanley (1,6), PhD

@@

(1) Department of Nutritional Sciences, University of Toronto, Toronto, Ontario, Canada.
(2) Centre for Studies in Family Medicine, University of Western Ontario, London, Ontario, Canada.
(3) Division of Endocrinology, University of Toronto, Toronto, ON, Canada.
(4) Lunenfeld Tanenbaum Research Institute, Mount Sinai Hospital, Toronto, Ontario, Canada.
(5) Department of Physiology, University of Toronto, Toronto, Ontario, Canada.
(6) Dalla Lana School of Public Health, University of Toronto, Toronto, Ontario, Canada.

@@

- Name: Anthony J. Hanley
- Current address:  
    Department of Nutritional Sciences  
    Faculty of Medicine  
    University of Toronto  
    FitzGerald Building, 150 College Street, Room 341  
    Toronto, ON, Canada, M5S 3E2
- Phone number: 416.978.3616
- Fax number:
- Email: anthony.hanley@utoronto.ca

Word count: 3543 / 4000

@@

# Abstract
250 / 250 words

Objective: Our aim was to determine the longitudinal associations of individual
non-esterified fatty acid (NEFA) species on changes in insulin sensitivity and
beta-cell function over 6 years in a cohort of individuals who are at-risk for
diabetes.

Research Design and Methods: In the PROMISE cohort, 477 participants had serum
NEFA measured at the baseline visit and who had completed an OGTT at 3 time
points over 6 years. Outcome variables were calculated using the OGTT values.
Insulin sensitivity was assessed using 1/HOMA-IR and the Matsuda index (ISI),
while beta-cell function was assessed using the Insulinogenic index over HOMA-IR
(IGI/IR) and the Insulin Secretion-Sensitivity Index-2 (ISSI-2). Generalized
estimating equations was used, adjusting for time, waist, sex, ethnicity, age,
ALT, alcohol intake, physical activity, and family history of diabetes. NEFA
were analyzed as both a concentration and as a proportion of the total fraction.

Results: Insulin sensitivity and beta-cell function declined by 14-27% while
BMI did not change over the 6-years. Total NEFA, 16:0, 18:1n-9, and 18:2n-6 as
concentrations had negative associations with IGI/IR and ISSI-2. These four NEFA
variables had similar magnitudes of association, predicting between 4-8% lower
IGI/IR and ISSI-2 over the 6 years. No associations were seen for 1/HOMA-IR
or ISI, nor were any associations seen for the mol% data.

Conclusions: Total NEFA was a strong predictor for lower beta-cell function over
6 years. Our results suggest that association with beta-cell function is due to
the absolute size of the serum NEFA fraction, rather than the specific lipid
composition.

@@

TODO:
- Bootstrap for the longitudinal table data... like instead of an ANOVA/GEE for
the p-value...?
- What about adiponectin?
- Precise followup time for each participant?
- [X] Interaction by sex or ethnicity
- [X] Interaction by time
- [X] Sensitivity analysis with TG
- [X] PLS-DA with dysglycemia (for anyone who got it over the years).
- [X] Replace TAG with BaseTAG in the QIC (or use both)
- [X] Fix time axis labels on LCMM plot
- Describe LCMM and PLS-DA better in plots

@@

stab_dist <- stab('nefaDist', 'Raw concentration values (nmol/mL) of each non-esterified fatty acid in the PROMISE cohort at the baseline visit (2004-2006). Values are presented as mean (SD).')
sfig_corr <- sfig('corr', 'Pearson correlation heatmap of non-esterified fatty acids (nmol/mL) and basic PROMISE participant characteristics for the baseline visit (2004-2006). Darkness of the colour indicates the magnitude of the correlation, with blue indicating positive and orange indicating negative correlations.')
sfig_gee <- sfig('gee', 'Unadjusted generalized estimating equation modeling of the longitudinal association of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort. Models are only adjusted for time. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, presented as the dot size.')

@@

stab_gee_unadj <- stab('gee_unadj', 'Unadjusted generalized estimating equations models of the longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.  Models are only adjusted for time. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values were adjusted for the BH false discovery rate, indicated by asterisk.')
sfig_lcmm <- sfig('lcmm', 'Latent class mixed model (LCMM) analysis to identify individual classes of trajectories for log(ISSI-2) over the 6 years in the PROMISE cohort. LCMM is a technique that identifies groups of participants that share a similar underlying trajectory in beta-cell function over the 6 years (e.g. no change compared to declines in ISSI-2). Red lines indicate individuals with a high beta-cell function who stayed high, green represents those in the middle, and blue represents those who had the lowest beta-cell function.')
sfig_plsda_groups <- sfig('plsda_groups', 'Clustering of extracted components from the partial least squares discriminant analysis (PLS-DA) on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006). PLS-DA is a multivariate technique that The percent explained variance of each component is shown in brackets on each axis. Red, green, and blue lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis.')

@@

sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', 'Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis. Blue lines indicate dysglycemia conversion or maintanence and red lines indicate no dysglycemia status.')

@@

Total non-esterified fatty acids (NEFA) have been well documented to influence 
the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that 
exposure to high concentrations of NEFA can induce insulin resistance in 
insulin-sensitive tissues such as muscle and liver, and in addition can impair 
pancreatic beta-cell production of insulin [@Rachek2014a;@Cnop2008a].
Epidemiological and clinical studies have reported concordant findings, showing
in particular that elevated total plasma NEFA associates with an increased risk
for incident type 2 diabetes [@Daniele2014a;@Salgin2012a].

Much of the previous experimental work on the role of NEFA in type 2 diabetes
utilized individual lipid species such as palmitic acid (16:0) or oleic acid
(18:1n-9), or alternatively used specific oils such as soybean oil, which is
high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the
exposure to characterize the impact of total NEFA. However, fatty acids comprise
multiple classes of molecules with diverse physiological functions, and few 
studies have compared the effects of a broader spectrum of fatty acids. Notably,
one study suggests that eicosapentaenoic acid (20:5n-3) can protect against the
lipotoxic effect of palmitic acid in the beta-cells [@Xiao2006].

Despite a sizable literature studying the role of total NEFA concentration in 
diabetes, there are important gaps in the research. The majority of previous 
studies have used animal models or cell lines [@Szendroedi2012a;@Kehlenbrink2012a], 
have been short term human infusion trials [@Daniele2014a;@Liang2013a], or have
been epidemiological studies that only looked at total NEFA and not individual
species [@Miller2012a;@Johns2014a]. To date, there have
been no longitudinal studies examining the role of the composition of the NEFA
blood lipid fraction on the pathogenesis of diabetes, which is critical to
understanding associations at a more granular level given protracted natural
history of diabetes and the growing appreciation of the divergent effects of
individual lipid species. Therefore, our research objective was to examine the 
association of serum NEFA composition on changes over time in insulin
sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
that higher palmitic acid and lower polyunsaturated fatty acids such as
eicosapentaenoic acid would associate with declining insulin sensitivity and
beta-cell function over 6 years.

@@

Participants from London and Toronto, Canada, were recruited into the 
Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
for recruitment into PROMISE required having one or more risk factors for type 2
diabetes mellitus, including obesity, hypertension, family history of diabetes, 
and/or a history of gestational diabetes or birth of a macrosomic infant. 
Participants aged 30 years and older (n=736) attended the baseline visit
between 2004-2006. Follow-up examinations in this cohort occur every three years, with three
examination visits completed to date (2004-2006, 2007-2009, and 2010-2013).  The
current study used data on participants who did not have diabetes at baseline,
who returned for one or more follow-up examinations, and who had samples
available for fatty acid measurements (n=477). A diagram of the sample size at
each visit is shown in `r cite_sf('consort')`. At each examination,
participants undergo metabolic characterization, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics. Research ethics approval
was obtained from Mount Sinai Hospital and the University of Western Ontario,
and all participants provided written informed consent.
Data collection methods were standardized across the 2 centres and research
nurses were centrally trained.

@@

At each examination, an 8-12 hour fasting blood sample was draw from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and triglycerides (TG) were measured using Roche
Modular's enzymatic colorimetric tests (Mississauga, ON).  Both specific insulin
and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
Basel, Switzerland) immunoassay analyzer and electrochemiluminescence
immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
and the Des 31,32 circulating split form (Linco Res. Inc), and has a CV of 9.3%.
Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
Diagnostics) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
performed at the Banting and Best Diabetes Centre Core Lab at Mt Sinai Hospital.
Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and diabetes
were categorized using the 2006 WHO criteria [@WHO2006].

NEFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of heptadecanoic acid (17:0) was added to the serum as an internal 
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction (NEFA, cholesteryl ester, phospholipid,
and triacylglycerol) was isolated using thin layer chromatography; each fraction was
visualized under UV light after lightly spraying with 8-anilino-1-naphthalene
sulfonic acid (0.1% wt/vol) and then converted to fatty acid methyl esters with
14% boron trifluoride in methanol at 100°C for 1 h. Fatty acid methyl esters
were separated and quantified using a Varian-430 gas chromatograph (Varian, Lake
Forest, CA, USA) equipped with a Varian Factor Four capillary column and a flame
ionization detector, which were injected in splitless mode. Fatty acid
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. 
There were 22 fatty acids measured in the NEFA fraction. Given the
diverse biology, as well as the complexity of the analyses, findings for other lipid
fractions in this cohort are reported separately (see {{pl and ce paper}} for the
analysis of the phospholipid and cholesteryl ester fractions).

@@

Insulin sensitivity and beta-cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using 1/HOMA-IR (1
divided by HOMA-IR) [@Matthews1985] and the Insulin Sensitivity Index (ISI)
[@Matsuda1999]. HOMA-IR largely reflects hepatic insulin resistance, while ISI
reflects whole-body insulin sensitivity [@AbdulGhani2007].  Beta-cell function
was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR)
and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
is a measure of the first phase insulin secretion while ISSI-2 is analogous to
the disposition index (but using OGTT values). Each index has been validated
against gold standard measures [@Matthews1985; @Matsuda1999; @Retnakaran2009].

@@

Height, weight, and waist circumference (WC) were measured at all clinic 
examinations using standard procedures.  WC was measured at the natural waist, 
defined as the narrowest part of the torso between the umbilicus and the xiphoid
process. BMI was calculated by dividing weight (kg) by height (m) squared.
Sociodemographic information, including age, sex, and ethnicity, were determined
using questionnaires administered at each examination. In the lifestyle
questionnaire, physical activity was determined using a version of the
Modifiable Activity Questionnaire (MAQ) [@Kriska1990].  The MAQ collects 
information on leisure and occupational activity, including intensity, 
frequency, and duration, over the past year.  Each reported activity from the 
MAQ was weighted by its metabolic intensity allowing for the estimation of 
MET-hours per week.

@@

The primary outcome variables for this analysis were 
1/HOMA-IR, ISI, IGI/IR, and ISSI-2; outcome variables were log-transformed for
the statistical modeling. The primary predictor variables for this analysis were
22 individual NEFA species as mole percent (mol%) of the total fraction and as
concentration (nmol/mL). Pearson correlation coefficients were computed to
assess the relationships of individual NEFA with other continuous variables.

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design 
an auto-regressive (AR1) correlation matrix was chosen for the GEE models, 
though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
is well suited to longitudinal cohort studies given it's capacity to handle
missed visits. The predictor variables and continuous covariates were scaled
(mean centered and standardized). The NEFA variables were classified as
*time-independent* (held constant) as they were measured only at the baseline
visit, while the outcome variables and covariates were set as *time-dependent*.
No imputation was conducted on missing values.

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
(QIC). `r cite_st('qic')` shows the covariates compared using QIC and 
`r cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model 
we selected differed between insulin sensitivity and beta-cell function models. 
For insulin sensitivity, we adjusted for time, sex, ethnicity, baseline age, WC,
ALT, MET, alcohol intake, and family history of diabetes (M6). For beta-cell
function, the last few models (M5-M8) were comparable in performance, so we
report results from the simplest model (M8), adjusting for time, sex, ethnicity,
baseline age, WC, ALT,
and family history of diabetes. After scaling, log-transforming, and
exponentiating the GEE estimates, the GEE results are interpreted as an expected
percent difference in the outcome variable for every SD increase in the
predictor variable given the covariates are held constant. While we a priori
decided to test for and report on the time by predictor interaction results, we
also assessed the overall fit of models that included a time interaction, which
ultimately did not improve the overall fit of our models in terms of QIC
(M6.1 and M8.1 as seen
in `r cite_st('qic')`) and we therefore did not include them in the main effect
analysis. Given that TG is a risk factor
for diabetes and that NEFA contribute to TG production, TG may act as a mediator
between NEFA and the outcomes. To determine the role of TG in the
association between NEFA and the outcomes, TG was included in the GEE model as a
sensitivity analysis. Lastly for the GEE models, we tested for an interaction
with sex or ethnicity by the predictor term for each outcome variable.

While GEE accounts for the longitudinal design of the data, this approach is
limited in that it cannot analyze the inherent multivariate nature of the
composition of the NEFA fraction. As such, after running the GEE models for the
longitudinal analysis, partial least squares discriminant analysis (PLS-DA) was
used to identify the patterns of NEFA composition against beta-cell function
(using ISSI-2, which had most of the associations in the GEE models) and for
those who converted to or maintained dysglycemia status (either IFG, IGT, or DM)
over the 6 years. Since PLS-DA cannot handle longitudinal data, latent class
mixed models (LCMM) were used to extract latent trajectories of beta-cell
function (ISSI-2) over the 6 years.

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE <!--[@Hoejsgaard2006a]-->, caret
`r packageVersion('caret')` for PLS-DA, and lcmm `r packageVersion('lcmm')` for
LCMM. The R code for this manuscript is available at {{link}}. Results were
statistically significance at p<0.05, after adjusting for multiple testing using
the Benjamini-Hochberg False Discovery Rate [@Benjamini1995a].

@@

gee_results_unadj <- dplyr::bind_rows(
    analyze_gee(gee_df, outcomes, ne_pct, 'VN', 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE)
    ) %>% 
    tidy_gee_results()
```

```{r lcmmPls, cache=TRUE, results='hide'}
# include also the 3-yr and 6-yr data for training/testing.
lcmm_results <- analyze_lcmm(ds)
lcmm_data <- combine_lcmm_data(ds, lcmm_results)
plsda_results <- analyze_plsda(lcmm_data)
plsda_dysgly <- dysgly_data %>% 
    mutate(ConvertDysgly = factor(ConvertDysgly, levels = c(0, 1), labels = c('No', 'Yes'))) %>% 
    analyze_plsda('ConvertDysgly')
```

@@

For this study, there were 
`r sex.npct$npct` females and `r ethn.npct$npct` had European-ancestry.
The average age of the participants was `r aide::average(ds0yr$Age)` years 
(`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
subset of the cohort, `r fhd.npct$npct`, had a family history of diabetes.
The primary outcome variables had a significant median decline between 
`r percent.change.outcomes$chg` (p`r percent.change.outcomes$p`) over the 6
years in this analysis of the PROMISE cohort (n=`r percent.change.outcomes$n`).
Consistent with this decline, there were `r dm.convert`
participants who developed diabetes while `r predm.convert` participants either
converted to or maintained IFG or IGT status over the 6 years.
